Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-09 article/article.Rmd
Commit 2024-09-09 21:18

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
# footnote:
#   - code: 1
#     text: "This is the first author footnote."
#   - code: 2
#     text: "Another author footnote."
# abstract: |
#   In this study we investigate the characteristics of global horizontal irradiance
#   \(200\,\text{W}/\text{m}^{2}\). The most active period of these extreme events is
#   spring -- early summer with a secondary peak in autumn.

@@

# Abstract {.unnumbered}

@@

In this study, we investigate the characteristics of global horizontal irradiance
enhancement events induced by clouds over Thessaloniki for the period 1994 -- 2023
using data recorded every one minute.  To our knowledge, there isn't a similar study
with such a long time series.  We identified the cloud enhancement (CE) events by
creating an appropriate cloud-free irradiance reference using a radiative transfer
model and aerosol optical depth data from a collocated Cimel sun photometer and a
Brewer spectrophotometer. We found a trend in CE events of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.N",slope],0))`\pm
`r round(2*yearlyCE[var=="GLB_diff.N",slope.sd],0)`\,\text{cases}/\text{year}$,
and a trend in the corresponding irradiation of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.sum",slope],1))`\pm
`r round(2*yearlyCE[var=="GLB_diff.sum",slope.sd],1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June. CE events with duration
longer than 10\nobreakspace{}min are very rare
($<`r ceiling(100*ST_G0[,sum(GLB_ench.N>10)/.N])`\,\%$), with exceptions lasting
over an hour and up to 140 minutes.  Finally, we have detected enhancements above the
total solar irradiance at the top of the atmosphere for the same solar zenith angle
of up to
$204\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$40\,\text{W}/\text{m}^{2}$. Most of these extreme events occur in spring -- early
summer, with a secondary peak in autumn.

@@

We must note that the excess irradiation from the CE events that is received at the
surface does not affect the balance of the energy budget in the atmosphere. The
incoming solar radiation in the region is not increased by the OI but is rather
redistributed at the specific location as asdditional scattered radiation.  This is
also depicted by the ECE irradiance values, which exceed the equivalent cloud-free
irradiance by a significant amount.

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the number of occurrences of the CE events in percent, calculated
relative to the available one-minute data in each month.  Although CE events are
present throughout the year, showing a clear seasonal cycle, the most active months
are May and June, with their median representing about $3\,\%$ of the data.  The
least frequent events (about $1\,\%$) occur in August and during winter (December --
February).  This seasonality is a combined effect of different factors, among them
the types of clouds, their frequency of occurrence, the seasonally varying relative
position of the sun, as well as the local landscape characteristics that may
influence the formation of the clouds.  For example, during May and June the frequent
formation of cumulus clouds in the area leads to a significant increase in the CE
events.  However, the lack of detailed data in the area of cloud characteristics does
not allow further analysis. Finally, the interannual variability of the monthly CE
events is quite high as manifested by the size of the boxes and the large monthly
extremes, especially in the summer. Investigation of the causes of this variability
is beyond the scope of this study. It could possibly be related to the cloud cover
variability in winter and autumn over southern Europe, associated with the North
Atlantic Oscillation (NAO) circulation [@Chiacchio2010], or to the observed
decreasing trend in cloud cover as a result of global warming [e.g.,
@SanchezLorenzo2017].

@@

The distribution of the number of CE events as a function of OI is shown in
Figure\nobreakspace{}\ref{fig:ovir-distribution}.  There is an inverse relationship
between the frequency of CE events and OI with an exponential-like decline. This is
expected, as the stronger the CE events are, the rarer the conditions favouring the
occurrence of CE events.
For the majority (over $`r round(some01$Enh_pc_below_mean,0)`\,\%$) of
the CE events the OI is below the long-term average of
$`r round(some01$Enh_OI_AVG,1)`\,\text{W}/\text{m}^2$, while
$`r round(some01$Enh_pc_below_100,1)`\,\%$ of the events correspond to OI larger than
$100\,\text{W}/\text{m}^2$ and up to the highest value of
$`r round(some01$Max_OI,1)`\,\text{W}/\text{m}^2$.
This distribution is indicative of the magnitude and the probability of the expected
CE events over Thessaloniki.  Similar distribution of CE events, albeit with larger
OI values, has been reported by @Vamvakas2020, for the city of Patras. This site is
located $2.5^\circ$ south of Thessaloniki and is exposed to air masses coming mainly
from the eastern Mediterranean, resulting in different cloud patterns, that may
affect the characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several minutes or even
more than an hour. To study the characteristics of these consecutive events, we
grouped them into bins of increasing duration in steps of one minute.  We have
identified `r ST_G0[, .N]` groups of CE events in the whole period of study, where
the group with the longest duration of
`r ST_G0[, max(GLB_diff.N)]` minutes occurred on
`r strftime(ST_G0[which.max(GLB_diff.N), Date], "%d\\ %B %Y")`
with SZA ranging between
$`r round(ST_G0[which.max(GLB_diff.N), SZA.min],1)`^\circ$ and
$`r round(ST_G0[which.max(GLB_diff.N), SZA.max],1)`^\circ$.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the frequency
distribution of the CE events according to their duration.  We conclude that although
some groups of events last for more than an hour, about
$`r round(100*ST_G0[,sum(GLB_ench.N<=5)/.N],0)`\,\%$ have a duration of
equal or less than 5 minutes.

@@

An aspect of the CE events that is commonly reported and has some significance on the
solar energy production infrastructure are the extreme CE (ECE) events, where the
measured GHI exceeds the expected irradiance at the TOA at the same SZA
(Equation\nobreakspace{}\ref{eq:ECE}).  In analogy to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences}, we show in 
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE} the number of ECE
events relative to the total number of one-min observations in each month. Judging by
the distribution of the monthly medians, the most active months in ECE events are
January -- March and November and the least active are July and August. The highest
interannual variability in the 30-year record occurs from October to April. In the
summer period, the fraction of ECE events is very small, with almost zero variability
in July and August.  Although the most active months for CE events are May and June
(Figure\nobreakspace{}\ref{fig:relative-month-occurrences}), these months behave
differently for the ECE events, suggesting that the cumulus clouds that usually cause
enhancements of irradiance are incapable to produce extreme events, possibly because
of their size and structure, in conjunction with the smaller SZAs in the summer.

@@

By creating a cloud-free approximation representing the long- and short-term
variability of the expected cloud-free GHI, we were able to identify cases of CE
events in Thessaloniki for the period 1994 -- 2023.  After analyzing the CE cases, we
found an increase of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $79\,\%$ of them is equal
or bellow $5$ minutes.

We have observed ECE cases with GHI exceeding the TOA TSI on horizontal plane by more
than $200\,\text{W}/\text{m}^{2}$ while for $75\,\%$ of the cases the excess
irradiance (relative to TOA TSI on horizontal plane) is below
$40\,\text{W}/\text{m}^{2}$. The climatological characteristics of the ECE events
showed that the most active months are spread over half of the year, and particularly in
the periods March -- June and September -- October.  The magnitude of the ECE events
identified in Thessaloniki events does not exceed the values reported for sites with
more favourable conditions for the phenomenon [e.g., @Cordero2023]. Some of the
characteristics of CE and ECE events we analyzed have strong similarities with the
results of @Vamvakas2020 for the city of Patras, south of Thessaloniki, albeit with
differences in the magnitude of OI.

An interpretation of the CE trends shows that the interaction of GHI with the clouds,
through this 30-year period, is a dynamic phenomenon that needs further
investigation. For such studies, it would be essential to have more detailed
information on cloud characteristics, especially in order to investigate possible
associations of the observed trends with changes in climate.

>>>>

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
# footnote:
#   - code: 1
#     text: "This is the first author footnote."
#   - code: 2
#     text: "Another author footnote."
abstract: |
  In this study, we investigate the characteristics of global horizontal
  irradiance enhancement events induced by clouds over Thessaloniki for
  the period 1994 -- 2023 using data recorded every one minute. To our
  knowledge, there isn't a similar study with such a long time series. We
  identified the cloud enhancement (CE) events by creating an appropriate
  cloud-free irradiance reference using a radiative transfer model and
  aerosol optical depth data from a collocated Cimel sun photometer and a
  Brewer spectrophotometer. We found a trend in CE events of
  \(+112\pm 35\,\text{cases}/\text{year}\), and a trend in the
  corresponding irradiation of \(+329.9\pm 112\,\text{kJ}/\text{year}\).
  The peak of the CE events was observed during May and June. CE events
  with duration longer than 10\nobreakspace{}min are very rare
  (\(<8\,\%\)), with exceptions lasting over an hour and up to 140
  minutes. Finally, we have detected enhancements above the total solar
  irradiance at the top of the atmosphere for the same solar zenith angle
  of up to \(204\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the
  cases below \(40\,\text{W}/\text{m}^{2}\). Most of these extreme events
  occur in spring -- early summer, with a secondary peak in autumn.

@@

<!--
# Abstract {.unnumbered}

@@

In this study, we investigate the characteristics of global horizontal irradiance
enhancement events induced by clouds over Thessaloniki for the period 1994 -- 2023
using data recorded every one minute.  To our knowledge, there isn't a similar study
with such a long time series.  We identified the cloud enhancement (CE) events by
creating an appropriate cloud-free irradiance reference using a radiative transfer
model and aerosol optical depth data from a collocated Cimel sun photometer and a
Brewer spectrophotometer. We found a trend in CE events of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.N",slope],0))`\pm
`r round(2*yearlyCE[var=="GLB_diff.N",slope.sd],0)`\,\text{cases}/\text{year}$,
and a trend in the corresponding irradiation of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.sum",slope],1))`\pm
`r round(2*yearlyCE[var=="GLB_diff.sum",slope.sd],1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June. CE events with duration
longer than 10\nobreakspace{}min are very rare
($<`r ceiling(100*ST_G0[,sum(GLB_ench.N>10)/.N])`\,\%$), with exceptions lasting
over an hour and up to 140 minutes.  Finally, we have detected enhancements above the
total solar irradiance at the top of the atmosphere for the same solar zenith angle
of up to
$204\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$40\,\text{W}/\text{m}^{2}$. Most of these extreme events occur in spring -- early
summer, with a secondary peak in autumn.
-->

@@

We must note that the excess irradiation from the CE events that is received at the
surface does not affect the balance of the energy budget in the atmosphere. The
incoming solar radiation in the region is not increased by the OI but is rather
redistributed at the specific location as additional scattered radiation.  This is
also depicted by the ECE irradiance values, which exceed the equivalent cloud-free
irradiance by a significant amount.

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the number of occurrences of the CE events in percent, calculated
relative to the available one-minute data in each month.  Although CE events are
present throughout the year, showing a clear seasonal cycle, the most active months
are May and June, with their median representing about $3\,\%$ of the data.  The
least frequent events (about $1\,\%$) occur in August and during winter (December --
February).  This seasonality is a combined effect of different factors, among them
the types of clouds, their frequency of occurrence, the seasonally varying relative
position of the sun, as well as the local landscape characteristics that may
influence the formation of the clouds.  For example, during May and June the frequent
formation of cumulus clouds in the area leads to a significant increase in the CE
events.  However, the lack of detailed data on cloud characteristics in the area does
not allow further analysis. Finally, the interannual variability of the monthly CE
events is quite high as manifested by the size of the boxes and the large monthly
extremes, especially in the summer. Investigation of the causes of this variability
is beyond the scope of this study. It could possibly be related to the cloud cover
variability in winter and autumn over southern Europe, associated with the North
Atlantic Oscillation (NAO) circulation [@Chiacchio2010], or to the observed
decreasing trend in cloud cover as a result of global warming [e.g.,
@SanchezLorenzo2017].

@@

The distribution of the number of CE events as a function of OI is shown in
Figure\nobreakspace{}\ref{fig:ovir-distribution}.  There is an inverse relationship
between the frequency of CE events and OI with an exponential-like decline. This is
expected, as the stronger the CE events are, the rarer the conditions favoring the
occurrence of CE events.
For the majority (over $`r round(some01$Enh_pc_below_mean,0)`\,\%$) of
the CE events the OI is below the long-term average of
$`r round(some01$Enh_OI_AVG,1)`\,\text{W}/\text{m}^2$, while
$`r round(some01$Enh_pc_below_100,1)`\,\%$ of the events correspond to OI larger than
$100\,\text{W}/\text{m}^2$ and up to the highest value of
$`r round(some01$Max_OI,1)`\,\text{W}/\text{m}^2$.
This distribution is indicative of the magnitude and the probability of the expected
CE events over Thessaloniki.  Similar distribution of CE events, albeit with larger
OI values, has been reported by @Vamvakas2020, for the city of Patras. This site is
located $2.5^\circ$ south of Thessaloniki and is exposed to air masses coming mainly
from the eastern Mediterranean, resulting in different cloud patterns that may
affect the characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several minutes or even
more than an hour. To study the characteristics of these consecutive events, we
grouped them into bins of increasing duration in steps of one minute.  We have
identified `r ST_G0[, .N]` groups of CE events in the whole period of study, where
the group with the longest duration of
`r ST_G0[, max(GLB_diff.N)]` minutes occurred on
`r strftime(ST_G0[which.max(GLB_diff.N), Date], "%d\\ %B %Y")`
with SZA ranging between
$`r round(ST_G0[which.max(GLB_diff.N), SZA.min],1)`^\circ$ and
$`r round(ST_G0[which.max(GLB_diff.N), SZA.max],1)`^\circ$.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the frequency
distribution of the CE events according to their duration.  We conclude that although
some groups of events last for more than an hour, about
$`r round(100*ST_G0[,sum(GLB_ench.N<=5)/.N],0)`\,\%$ have a duration of
equal or less than 5 minutes, likely due to the movements of clouds causing frequent
blocking of the direct irradiance, which reduce the duration of the CE events.

@@

An aspect of the CE events that is commonly reported and has some significance on the
solar energy production infrastructure are the extreme CE (ECE) events, where the
measured GHI exceeds the expected irradiance at the TOA at the same SZA
(Equation\nobreakspace{}\ref{eq:ECE}).  In analogy to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences}, we show in 
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE} the number of ECE
events relative to the available number of one-min observations in each month. Judging by
the distribution of the monthly medians, the most active months in ECE events are
January -- March and November and the least active are July and August. The highest
interannual variability in the 30-year record occurs from October to April. In the
summer period, the fraction of ECE events is very small, with almost zero variability
in July and August.  Although the most active months for CE events are May and June
(Figure\nobreakspace{}\ref{fig:relative-month-occurrences}), these months behave
differently for the ECE events, suggesting that the cumulus clouds that usually cause
enhancements of irradiance are incapable to produce extreme events, possibly because
of their size and structure, in conjunction with the smaller SZAs in the summer.  In
contrast, during winter the Sun is lower in the sky and clouds are frequently less
dense allowing a large fraction of radiation to be scattered forward and enhance the
irradiance at the surface. These conclusions are based on manual inspection of many
sky-camera images during the summer and winter periods. An automated procedure of
checking the sky-camera images for such events would greatly improve our ability to
classify the CE and ECE events and help in the attribution of the causes of the
events.

@@

By creating a cloud-free approximation representing the long- and short-term
variability of the expected cloud-free GHI, we were able to identify cases of CE and
ECE events in Thessaloniki for the period 1994 -- 2023.  After analyzing the CE
cases, we found an increasing tendency of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $79\,\%$ of them is equal
or bellow $5$ minutes.

We have observed ECE cases with GHI exceeding the TOA TSI on a horizontal plane by
more than $200\,\text{W}/\text{m}^{2}$ while for $75\,\%$ of the cases the excess
irradiance (relative to TOA TSI on a horizontal plane) is below
$40\,\text{W}/\text{m}^{2}$. The climatological characteristics of the ECE events
showed that the most active months are spread over half of the year, and particularly
in the months January -- March and November, while ECE events are practically absent
in July and August.  The magnitude of the ECE events identified in Thessaloniki does
not exceed the values reported for sites with more favorable conditions for the
phenomenon [e.g., @Cordero2023]. Some of the characteristics of CE and ECE events we
analyzed have strong similarities with the results of @Vamvakas2020 for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude of OI.

The investigation of the CE trends in Thessaloniki through this 30-year period shows
that the interaction of GHI with clouds is a complex phenomenon that merits further
attention. For such studies, it would be essential to have more detailed information
on cloud characteristics, especially in order to investigate possible associations of
the observed trends with changes in climate.  Finally, automated procedures to
analyze the sky-camera images would greatly improve our knowledge about the
generation of these events.
